---
title: "Index of Multiple Deprivation"
description: |
  The measure of relative deprivation in small areas in England called lower-layer super output areas
author:
  - name: Zoë Turner
    url: {https://twitter.com/Letxuga007}
date: 05-14-2021
output:
  distill::distill_article:
    self_contained: false
categories:
  - Resources    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, results = 'asis')
library(dplyr)
library(knitr)
```

# Index of Multiple Deprivation (2019)

IMD is very useful for categorising the area a person lives in for deprivation. This information/links and code only cover England and deprivation is scored in relation to all the areas within England using the IMD (2019). This means that they have been ordered by the deprivation score and then ranked (IMDRank).

* The common decile used is between 1 and 10 with 10 being the least deprived.

# Data warehouses

The data is often held in 3 tables:

* the postcodes of the data held (for example patients when in the healthcare sector)
* a lookup postcode table (like a directory of postcodes) from 

https://digital.nhs.uk/services/organisation-data-service/data-downloads/ods-postcode-files 

* the IMD data from 

Older version: http://www.gov.uk/government/statistics/english-indices-of-deprivation-2015 
https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019 

Select File 7 for the dataset.

Other data is available from this dataset including IDAOPI which relates to only older people. 

Note that the column headers change, in 2015 it was LADistrictCode2013 and in 2019 it is LADistrictCode2019. Also LADistrictName2013 has become LADistrictName2019.

## Watch for...

### 'Unknown' in the data warehouse

Data warehouse tables may include a row for 'unknown' (https://www.sqlchick.com/entries/2011/5/16/usage-of-unknown-member-rows-in-a-data-warehouse.html) and this may create a valid join across tables: unknown is entered as a postcode that links to the postcode table that in turn returns data from the IMD table. This may result in a value being returned from the IMD table that isn't valid like 0 which looks like it should be an IMD decile score, for example, but which is not.
 
### Postcode spaces 
 
Also, postcode lengths vary according to how many spaces there are between the two parts. In the UK the postcode format can be 3 parts and then 3 or 4 then 3:

NG1 1AA
NG26 1AA

Joining datasets is always better when the space between the postcode parts is removed. In SQL this can be:

REPLACE(postcode, ' ', '')

in R it can be 


```{r Postcode White Space, echo=TRUE}
postcode <- "NG16 1AA"
gsub(" ","",postcode)
```


### Partial postcodes

Partial postcodes will not give a sufficiently reliable IMD score.


# Example join code (SQL)

To get the IMD score the LSOA (Lower Super Output Area) code is required which is taken from the full postcode.

This code will not run and is dependent on the naming conventions of the SQL server. The column names of LSOA11 and LSOAcode2011 will have come from the data sources. PostCode_space will have been added to the table by the data warehouse administrator(s).

```{r SQL Example code, eval=FALSE}
SELECT Top 100 imd.*
FROM DIM_AI.PatientData AS p
LEFT JOIN DIM_AI.PostCodes AS pc ON p.PostCode = pc.PostCode_space												
LEFT JOIN DIM_AI.IMD AS i ON pc.PC.LSOA11 = i.LSOAcode2011
WHERE p.PostCode LIKE 'NG%'
```

# Creating quintiles

Some publicly available data is in quintiles for IMD to remove small identifiable numbers. 

To replicate that in local data the equation: floor((IMDDecile-1)/2) + 1 can be applied.

### Quintiles in SQL

```{r Quintile SQL code, eval=FALSE}
SELECT DISTINCT IMDDecile,
FLOOR((IMDDecile-1)/2) + 1 AS IMDQuintile
FROM DIM_AI.IMD
ORDER BY IMDDecile
```

### Quintiles in R

The following example Pubicly available data will not run download if this Rmarkdown script is run as the eval has been set to FALSE. Either change this to TRUE, remove eval=FALSE altogether or copy the code to another R script to run.

```{r ONS quintile example, eval=FALSE}
# IMD by Ethnicity by Region 2007-2013
# Latest: https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/adhocs/006134birthsbyethnicitysexregionandimdquintilebyfinancialyear2007to2013
download.file("https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/adhocs/006134birthsbyethnicitysexregionandimdquintilebyfinancialyear2007to2013",
              destfile = "regionbirthsbyethnicityIMD20072013.xls",
              method = "wininet", #use "curl" for OS X / Linux, "wininet" for Windows
              mode = "wb") #wb means "write binary"
```

Applying the formula:

```{r quintile}
library(tidyverse)
# Generate a dataset
df <- structure(list(IMDDecile = c(0L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 
8L, 9L, 10L, NA)), row.names = c(NA, -12L), class = c("tbl_df", 
"tbl", "data.frame"))
# Make the 0 generated into NA
df_quintile <- df %>% 
  replace_na(list(IMDDecile = 0)) %>% 
  mutate(IMDQuintile = floor((IMDDecile-1)/2) + 1,
         IMDQuintile = as.character(IMDQuintile)
         )
```

# Creating local IMDs

In Nottingham/Nottinghamshire the differences between the LSOA areas is diminished when ranked against England as a whole, but when ranked locally, the variation is much more pronounced. Consequently, for the majority of our analysis the Trust approach is to use Nottinghamshire quintiles of deprivation / IMD.  There will be some instances where national quintiles (or deciles) will be required – namely any external analysis – however, these will be the exception. 

For note, when using local quintiles anyone with a non-local postcode will come back unmatched as well as those people with proxy postcodes used often to denote homeless status (ZZ...). 

## IMD in SQL

To create the local rankings in SQL the data needs to be restricted to the appropriate area, for example when joining to the Postcodes and restricting and then a windows partition applied to the data ROW_NUMBER() OVER(ORDER BY IMDRank) to create a new ranking score and NTILE(10) OVER (ORDER BY IMDRank) to create new deciles.

To replicate this in dplyr the code would be:

```{r localRank, eval=TRUE, echo=TRUE}
 # Data for Nottingham and Nottinghamshire taken from Postcodes and IMD. The original files are very large to download. 
# Note that there are columns in the Postcode data sample that don't exist in the source file. 
# These have been added locally but may be of use to others, such as CountyName and LocalAuthority_Name. # These are specifically added as Nottingham Local Authority is a Unitary Authority and appears in a different column to Nottinghamshire County's District Councils 
load("data/sampleDataNottingham.RData")
library(dplyr)
# Note on the join the two column names are different so are listed in the by = and they need to be in the correct order so the column from imdNottingham appears first.
localRanking <- imdNottingham %>% 
  inner_join(postcodeData %>% 
               select(LSOA11) %>% 
               group_by(LSOA11) %>% 
               slice(1), by = c("LSOAcode2011" = "LSOA11")) %>% 
  mutate(Notts_rank = row_number(IMDRank),
         Notts_decile = ntile(IMDRank, 10)) 
localRanking %>% 
  select(LSOAcode2011:IMDDecile) %>% 
  head(5) %>% 
  knitr::kable(format = "html")
```

# Other useful links

https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/464430/English_Index_of_Multiple_Deprivation_2015_-_Guidance.pdf 

https://fingertips.phe.org.uk/search/imd

http://dclgapps.communities.gov.uk/imd/idmap.html 

Technical report for 2019: https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/833951/IoD2019_Technical_Report.pdf 


# Referencing IMD in a paper or research
 
From a journal check to see how IMD is referenced in published papers, this was from a 2016 BMJ article that cites the Index in the references as:

6. Department for Communities and Local Government. English indices of deprivation 2015. 2015. https://www.gov.uk/government/statistics/ english-indices-of-deprivation-2015

(Taken from https://bmjopen.bmj.com/content/bmjopen/6/11/e012750.full.pdf)

Another paper from 2016 cites as:

16. Department for Communities and Local Government. English Indices of Deprivation 2015. Available online: http://www.gov.uk/government/statistics/english-indices-of-deprivation-2015 (accessed on 27 April 2016).

(taken from https://www.mdpi.com/1660-4601/13/8/750)

Looking at the Government page that lists the full text the library assistant said: "I would reference it from the Ministry of Housing, Communities and Local Government which would be more up to date for 2019 and with online references you should always put the date you accessed it. So I would suggest amending to the following:"

Ministry of Housing, Communities and Local Government. English Indices of Deprivation 2015. 2015. https://www.gov.uk/government/statistics/english-indices-of-deprivation-2015 (Accessed 4 June 2019)
