---
title: "CDU Data Science Team &mdash; Roadmap"
subtitle: "Branded slides in R"
author: "Chris Beeley"
institute: "Nottinghamshire Healthcare NHS Foundation Trust"
date: "March 2021"
output:
  xaringan::moon_reader:
    css: 
      - default
      - css/nhsr.css
      - css/nhsr-fonts.css
    lib_dir: libs
    seal: false
    self_contained: true
    nature:
      highlightStyle: googlecode
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
---
class: title-slide, left, bottom
```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
library(nhsrtheme)
library(xaringanExtra)
# set default options
opts_chunk$set(echo = FALSE,
               fig.width = 7.252,
               fig.height = 4,
               dpi = 300,
               dev.args = list(type = "cairo"),
               echo = TRUE,
               warning = FALSE,
               message = FALSE,
               error = FALSE)

# uncomment the following lines if you want to use the NHS-R theme colours by default
# scale_fill_continuous <- partial(scale_fill_nhs, discrete = FALSE)
# scale_fill_discrete <- partial(scale_fill_nhs, discrete = TRUE)
# scale_colour_continuous <- partial(scale_colour_nhs, discrete = FALSE)
# scale_colour_discrete <- partial(scale_colour_nhs, discrete = TRUE)

library(xaringanExtra)
xaringanExtra::use_share_again() # need to get the slide button on html view
```

```{r echo=FALSE}
xaringanExtra::use_logo(
  image_url = "img/nottshc.png",
  exclude_class = c("title-slide", "inverse", "hide-logo"),
  width = "300px",
  height = "200px",
  position = css_position(top = "0em", right = "1em")
)

```


# `r rmarkdown::metadata$title`
----
## **`r rmarkdown::metadata$subtitle`**
### `r rmarkdown::metadata$author` | `r rmarkdown::metadata$date`
</br>

---

# Themes 

</br>
* Making sense of Trust data and reporting 

* Making predictions and inferences with statistics and machine learning 

* Sharing code and knowledge 

---

# Activity

</br>

* Patient and staff experience 

* Routine outcomes

* Clinical need and workload prioritisation

* Forecasting

* Statistical process control

* R training

* Statistics and machine learning training and clinics

* Predicting Did Not Attend with machine learning

---

# Patient and staff experience 

.left-column[

]

.left-column[
## Done

* [R Shiny dashboard](http://suce.co.uk:8080/apps/SUCE/) summarising patient experience
* R Shiny [staff experience dashboards](https://github.com/ChrisBeeley/staff_survey)
* Automatic reports for number of responses and quality of experience
]

.left-column[
## Doing

* [Text mining](https://github.com/CDU-data-science-team/positive_about_change_text_mining) patient experience comments to tag by theme and sentiment
* Sharing text mining with five other provider trusts
]

.left-column[
## Considering

* General purpose machine learning system which can learn any tags for any dataset (with some human input)
* Producing dashboard/ reports for staff/ patient experience and sharing with other provider trusts
]

---

# Routine outcomes

.left-column[

]

.left-column[
## Done

* Summary of mortality
* Scoping of datasets and outcomes in mental health services
]

.left-column[
## Doing

* Analysis of HoNOS, referral and activity information, mortality, and staff patient experience
* Analysis of health inequalities

]

.left-column[
## Considering

* Analysis of system outcomes and activity from GPRCC
* Analysis of staff variables like sickness, turnover, and vacancy rate
* Analysis of risk and incidents

]

---

# Clinical need and </br> workload prioritisation

.left-column[

]

.left-column[
## Done

* Activity summaries (HealthCARD)
* Review of caseload rank score (from HealthCARD)

]

.left-column[
## Doing

* Produce and validate acuity score

]

.left-column[
## Considering

* Prediction of hospitalisation, service use, readmission, and clinical risk

]

---

# Forecasting

.left-column[

]

.left-column[
## Done

* Prediction of numbers of clinically safe beds in Nottingham University Hospitals

]

.left-column[
## Doing

* Predicting drug issues in pharmacies in Nottinghamshire Healthcare

]

.left-column[
## Considering

* ???

]

---

# Statistical Process Control Charts </br> (SPC)

.left-column[

]

.left-column[
## Done

* [Prototype SPC dashboard](https://github.com/CDU-data-science-team/healthcareSPC) (R/ Shiny)

]

.left-column[
## Doing

* Funding application to produce an SPC dashboard for the whole NHS

]

.left-column[
## Considering

* ???

]

---

# R Training

.left-column[

]

.left-column[
## Done

* [Delivered intro to R](https://nhsrcommunity.com/events/) to 100+ trainees around the country
* Developed and delivered[ intro to Shiny course](https://github.com/nhs-r-community/shiny-training) and for [R beginners](https://github.com/ChrisBeeley/shiny_beginners)
* Delivered ["R for SPSS users"](https://github.com/ChrisBeeley/r-for-spss) to 10 staff in the Trust

]

.left-column[
## Doing

* Funding application for Advanced Shiny and Introduction to Git training course

]

.left-column[
## Considering

* ???

]

---

# Statistics and machine learning </br> training and clinics

.left-column[

]

.left-column[
## Done

* Delivered Understanding Data courses to 50+ attendees
* Delivered statistics tutorials for 5 analysts
* Run ["Help with data"](https://cdu-data-science-team.github.io/team-blog/posts/2021-01-13-data-clinics-in-nottinghamshire-healthcare/) clinics for 3 teams in the trust

]

.left-column[
## Doing

* Working with Research and Evaluation to offer advice to their monthly clinics

]

.left-column[
## Considering

* Mentoring offered to analysts across the ICS
* Stats and ML tutorials for ICS analysts

]

---

# Predicting Did Not Attend </br> with machine learning

.left-column[

]

.left-column[
## Done

* Basic algorithm complete

]

.left-column[
## Doing

* Producing a dashboard to work with the data and handing the work over to Applied Information

]

.left-column[
## Considering

* Predict other healthcare events using machine learning (including free text)

]

---
class: inverse

name: acknowledgement

# Acknowledgments

Acknowledgements: the professional look of this presentation, using NHS and Nottinghamshire Healthcare NHS Foundation Trust colour branding, exists because of the amazing work of Silvia Canelón, details of the workshops she ran at the [NHS-R Community conference](https://spcanelon.github.io/xaringan-basics-and-beyond/index.html).

[`r icon::fontawesome("twitter")` @DataScienceNott](https://twitter.com/DataScienceNott)<br/>
[`r icon::fontawesome("github")` Clinical Development Unit Data Science Team](https://github.com/CDU-data-science-team)<br/>
[`r icon::fontawesome("paper-plane")` zoe.turner2@notthshc.nhs.uk](mailto:zoe.turner2@nottshc.nhs.uk)

