---
title: "Personal Access Tokens"
description: |
  Connecting to the GitHub to install packages from a private GitHub repository requires security 
  set ups and this blog details how to do it (and how not to do it). 
author:
  - name: Zoë Turner
    url: {https://twitter.com/Letxuga007}
date: 2021-12-17
output:
  distill::distill_article:
    self_contained: false
categories:
  - Packages
  - Security
preview: img/british-mineralogy-or-coloured-figures-p132.jpg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Installation

Because some of the packages we are using are (currently) private repositories it is not possible to build using just {remotes} as a Personal Access Token, specifically a GITHUB_PAT, is needed. The {usethis} package suggests using the local Git credential store but this wasn't compatible with some of our team's set ups so later in this blog are details on how to do it the "inadvisable" way to be used at your own discretion.

## Check you have a token

This part is the same for whatever method you use on your own computer as this section details how to get the GITHUB_PAT set up. To check for existing GITHUB_PAT use code:

`Sys.getenv("GITHUB_PAT")`

If nothing has been set up by you on GitHub this may return an empty string "".

## Set up a token

Firstly, go to https://github.com/settings/tokens which takes you to your own personal GitHub account settings where you will need to `Generate new token`. Give this a suitable name like `Repo access` and tick the `repo` group for full control of private repositories.

_Remember to copy the code that is generated as this cannot be viewed again_


# From the Git credential store

## Packages required

```{r eval=FALSE, echo=TRUE}

install.packages("usethis")
install.packages("gitcreds")
install.packages("remotes")

```


{usethis} suggests using the [local Git credential store](https://usethis.r-lib.org/articles/git-credentials.html) quite strongly!

> If you have previously set your GitHub PAT in .Renviron, stop doing that.

## Set up

```{r}
usethis::create_github_token()
```

GitHub will open and, if you are not already logged in, you will need to enter your (GitHub) password.

The page for GitHub tokens management is https://github.com/settings/tokens and you can do the same as the code
in this screen by selecting `Generate new token`. Give this a suitable name like `Package installation` and tick the `repo` group for full control of private repositories.

_Remember to copy the code that is generated as this cannot be viewed again_

Next set up a `GITHUB_PAT` in RStudio:

``` r
gitcreds::gitcreds_set()
```

You will then be prompted to enter the copied token (no quotes are required). If you've already got a token entered the following message will show:

```
-> Your current credentials for 'https://github.com':

  protocol: https
  host    : github.com
  username: PersonalAccessToken
  password: <-- hidden -->

-> What would you like to do? 

1: Keep these credentials
2: Replace these credentials
3: See the password / token

Selection: ```
Enter an item from the menu, or 0 to exit
Selection: 
```

Reference: https://usethis.r-lib.org/articles/articles/git-credentials.html

## Installing

The {usethis} documentation suggests using {pak} but this may not work on network drives/VPNs (there are a few issues that have been opened and quickly closed referring to these as potential problems) and so in the meantime 
combining {remotes} with {gitcreds} is a workaround:

``` r
# install.packages("remotes")
remotes::install_github("<name of GitHub account>/<name of repository>", 
    auth_token = gitcreds::gitcreds_get(use_cache = FALSE)$password)

```

# The absolutely not recommended method

## Packages required

```{r eval=FALSE, echo=TRUE}

install.packages("usethis")
install.packages("remotes")

```

If you are sure this is what you want to do then type in RStudio:

``` r
usethis::edit_r_environ()

```
In the file that opens, type into it (where YOUR-PAT is the generated code you've copied) and save the file:

``` r
GITHUB_PAT=YOUR-PAT
```

Restart R and run `Sys.getenv("GITHUB_PAT")` to check that the `""` has changed.

Reference: https://www.jumpingrivers.com/t/2019-user-git/03-githubpat.html#6

### Installing

Now, for as long as the token exists, the following code will install the package:

``` r
# install.packages("remotes")
remotes::install_github("<name of GitHub account>/<name of repository>", auth_token = Sys.getenv("GITHUB_PAT"))

```

# Building package

It is preferable to build from the repository rather than locally as this should be what everyone in the team has access to. To build and test any new functions use `Ctrl+Shift+L` from the {devtools} package to load locally for the session. 