---
title: "Using {usethis} to set up .gitignore"
description: |
  How to ensure that certain files cannot be accidentally committed to GitHub (or any other version controlled area).
author:
  - name: Zoë Turner
    url:
      https://twitter.com/letxuga007: {}
date: 2022-04-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - Open source
  - GitHub
preview: img/grundzüge_der_mathematischen_geographie_und_der_landkartenprojection.jpg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

# Use .gitignore

.gitignore is used to prevent committing certain files or folders to GitHub and can be set both locally in a project and globally. 

Using the {usethis} package a lot of the set up of these files can be done from the RStudio Console command line.

# Set up

These [slides](https://forwards.github.io/workshops/package-dev-modules/slides/02-setting-up-system/setting-up-system.html#17) from Forwards are for package development but are fantastic to follow to get your computer set up with GitHub.

Check the Git connections are set up with:

```{r}
usethis::git_sitrep()
```

If `Vaccinated: FALSE` which it is most likely to be if this is the first time {usethis} was used type:

```{r}
usethis::git_vaccinate()
```

This adds `.DS_Store`, `.Rproj.user`, `.Rdata`, `.Rhistory`, and `.httr-oauth` to the global `.gitignore` which can be checked with:

```{r}
usethis::edit_git_ignore(scope = "user")
```

This open file can be altered directly and saved.

## Changing project .gitignore

To view and edit directly:

```{r}
usethis::edit_git_ignore(scope = "project")
```

Or use the following to add individual files and folders like `secrets/`:

```{r}
usethis::use_git_ignore("secrets/")
```


# Potential issues with networks/VPNs

If `git_vaccinate` has and error starting with 

> libgit2::git_config_set_string

and then a reference to a lock to a private network drive (usually when off the VPN/Network) then type:

```{r}
usethis::edit_git_config()
```

and copy the following code, with <> changed as appropriate:

```
[user]
	email = <Email>
	name = <GitHub Account name>
[core]
	excludesFile = C:/Users/<own folder>/.gitignore
[init]
	defaultBranch = main

```

# Overwriting a global .gitignore

This was a question that came out of a team Code Review and we found the following [blog from Scott Radcliff](http://scottradcliff.com/how-to-overwrite-a-global-gitignore.html) with the answer that yes, it is possible to overwrite a global setting if this is ever required.

Not tested but examples using the `!` to exempt:

* for single files `!data.csv`
* for a group of files `!*.csv`
* for a subfolder `!secrets/not-secret/` from [Stackoverflow](https://stackoverflow.com/questions/3203228/git-ignore-exception)
* override all rules in global `!*` from [Stackoverflow](https://stackoverflow.com/questions/26678955/un-ignore-all-files-in-global-gitignore)

# .rda and .RData files

We also discussed `.rda` and `.Rdata` files which are save R data files. `gitvaccinate()` adds the `.Rdata` but doesn't mention `.rda` files so be aware of that.

The difference in the types of files are that `.Rdata` can store single or multiple R objects but `.rda` can only save single. 

It might be worthy of a blog/team review on its own but there are issues with saving multiple R objects to one place if they are updated regularly. It's very easy to not re-save/update an R object and affect the overall data workflow and so I, particularly as I've got into this very muddle, only save one R object per `.Rdata` file.
